---
title: "LinkOrgs Package Tutorial"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LinkOrgs Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[UTF-8]{inputenc}
---

## Introduction

This vignette demonstrates how to use the fastrerandomize package for generating and testing experimental designs in an agricultural study setting based on: 

* Burchardi, K.B., Gulesci, S., Lerva, B. and Sulaiman, M., 2019. Moral hazard: Experimental Evidence from Tenancy Contracts. The Quarterly Journal of Economics, 134(1), pp.281-347.

We will walk through:

* Generating synthetic covariate data
* Creating candidate randomizations
* Exploring the S3 methods for printing, summarizing, and plotting randomization objects
* Conducting a randomization-based inference test

Note: This tutorial assumes you have installed the fastrerandomize package, either from source or via GitHub:

```{r setup, eval=FALSE}
# If you haven't installed or set up the package:
# devtools::install_github("cjerzak/fastrerandomize-software/fastrerandomize")

# (Done once) Optionally build the JAX backend if needed
# fastrerandomize::build_backend()

# note that this vignette can be found like so: 
# vignette(package = "fastrerandomize")
```

## 1. Obtain Pre-treatment Covariate Data

For illustration, we use several pre-treatment covariates from the original experiment. 

```{r generate_covariates}
library(fastrerandomize)

# Obtain pre-treatment covariates 
data(QJEData, package = "fastrerandomize")
myCovariates <- c("children","married","hh_size","hh_sexrat")
QJEData <- QJEData[!is.na(rowSums(QJEData[,myCovariates])),]
X <- QJEData[,myCovariates]
  
# Analysis parameters
n_units   <- nrow(X)
n_treated <- round(nrow(X)/2)
```

## 2. Generate Randomizations

We now create our candidate randomizations using the function `generate_randomizations()` from fastrerandomize. Depending on the scale and complexity of your experiment, you may choose either exact enumeration ("exact") or Monte Carlo ("monte_carlo") randomization.

In this example, we'll use the exact approach, but note that `randomization_accept_prob` is set very low (0.0001) for demonstration, so that we can see some filtering in action.

```{r generate_randomizations}
RunMainAnalysis <- (!is.null(check_jax_availability()) )
# if FALSE, consider calling build_backend()

if(RunMainAnalysis){
CandRandomizations <- generate_randomizations(
  n_units = n_units,
  n_treated = n_treated,
  X = X,
  randomization_type = "monte_carlo", 
  max_draws = 100000L,
  batch_size = 1000L,
  randomization_accept_prob = 0.0001
)
}
```

## 4. S3 Methods: Print, Summary, and Plot

`generate_randomizations()` returns an S3 object of class `fastrerandomize_randomizations`. We can use its associated print, summary, and plot methods to inspect the candidate randomizations.

```{r s3_methods}
if(RunMainAnalysis){
# 4a. Print the object
print(CandRandomizations)

# 4b. Summary
summary(CandRandomizations)

# 4c. Plot the balance distribution
plot(CandRandomizations)
}
```

## 3. Randomization Test

Next, we showcase a randomization-based inference procedure using fastrerandomize.

### 3a. Setup Simulated Outcomes

We simulate an outcome Y that depends on our covariates X, the chosen randomization, and some true treatment effect τ. Here, we pick the first acceptable randomization from our set as our "observed" assignment, Wobs.

```{r setup_outcomes}
if(RunMainAnalysis){
CoefY <- rnorm(ncol(X))          # Coefficients for outcome model
Wobs  <- CandRandomizations$randomizations[1, ]  # use the 1st acceptable randomization
tau_true <- 1                     # true treatment effect
# Generate Y = X * Coef + W*tau + noise
Yobs <- as.numeric( as.matrix(X) %*% as.matrix(CoefY) + Wobs * tau_true + rnorm(n_units, sd = 0.1))
}
```

### 3b. Run the Randomization Test

We pass our observed treatment assignment (`obsW`), observed outcomes (`obsY`), and the full matrix of candidate randomizations (`candidate_randomizations`) to `randomization_test()`. This test:

* Computes the observed difference in means
* Permutes the treatment assignment across our candidate randomizations
* Estimates the p-value by comparing the observed statistic to its permutation distribution

```{r randomization_test}
if(RunMainAnalysis){
randomization_test_results <- randomization_test(
  obsW = Wobs,
  obsY = Yobs,
  candidate_randomizations = CandRandomizations$randomizations,
  findFI = TRUE
)

# Inspect results
print(randomization_test_results)
summary(randomization_test_results)
plot(randomization_test_results)
}
```

The `findFI = TRUE` flag further attempts to locate a fiducial interval (FI) for the treatment effect.

## Conclusion

This tutorial demonstrated a basic workflow with fastrerandomize:

* Generating synthetic or real covariate data
* Producing a set of acceptable treatment assignments via randomization
* Visualizing and summarizing these assignments
* Testing for treatment effects with a randomization-based approach

We encourage you to consult the package documentation for more advanced functionalities, including GPU-accelerated computations via JAX and flexible definitions of balance criteria in addition to what was shown here.